---
title: Effects of Car Transmission Types on Miles per Gallon Analysis using Linear
  Regression
output: pdf_document
geometry: margin=0.7in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
#load libraries and dataset
#install.packages("ggplot2", repos = "http://cran.us.r-project.org")
#install.packages("car", repos = "http://cran.us.r-project.org")
library("stats")
library('car')
library("ggplot2")
data("mtcars")
```
### Executive Summary
In this exploration, we look at the data from [Motors Trends](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html) magazine collected in 1974 for 32 car models. Our goal is to analyze the potential impact of car transmission types over miles per gallon efficiency by answering two questions:

- “Is an automatic or manual transmission better for MPG”
- "Quantify the MPG difference between automatic and manual transmissions"

From the data exploration and completed linear regression, we can conclude that manual transmission cars have a marginal impact over mpg efficiency but not strong enough to be seen as the sole variable predicting mpg efficiency. Our multivariate analysis has shown that instead, a combination of predictors such as: cylinder, weight and horse power including transmission help explain mpg efficiency over manual transmission alone.

##### Data Exploration and Analysis
The initial analysis will be based on a simple linear regression model with the two variables the question focuses on:

- `mpg`: Miles/(US) gallon, which will represent the outcome in this linear model.
- `am`: the car transmission. In our model, this will be our predictor variable, which has 2 possible values: 0 = automatic, 1 = manual. (See point A in the appendix.) . Therefore, transmission will be treated as a factor variable.

Looking further into the dataset (see point A in the appendix.) we can see several variables with levels, therefore, we'll proceed with a quick transformation of these variables into factors preemptively assuming we will need them for further exploration of our intital model. Said variables are: `cyl`, `gear` and `carb`, and `vs`.
```{r, echo= FALSE}
mtcars$am <- factor(mtcars$am)
mtcars$cyl <- factor(mtcars$cyl)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
mtcars$vs <- factor(mtcars$vs)
```

*I. Initial Model: MPG by transmission predictor*

We'll start by plotting mpg by transmission type, looking at the box plot (point B in the appendix), we can see that manual cars has a better impact over mpg.
```{r}
model.1 <- lm(mpg ~ am, data = mtcars)
summary(model.1)$coef
confint(model.1)
```
Digging deeper, using our initial linear `model.1`, we can see an increase in MPG for cars with manual transmissions of `r summary(model.1)$coef[2]` over automatic transmission. Looking as well at the low p-value `0.000285` and that the confidence interval does not include zero, we can reject a null hypothesis of no difference between manual vs. automatic transmission. However, looking closer at the model, its fit is pretty low as seen through an 
`r summary(model.1)$r.squared` r-squared value which doesn't seem sufficient to explain the mpg variability. 
It seems best considering the dataset includes 10 dimensions, to explore if other variables in addition to transmission `am` have an impact on mpg efficiency.

*Multivariate regression and model selection*

To figure out if other predictors do have impact over mpg, we'll start with a model that includes all captured variables in the dataset. 
We will then proceed with a *setwise model* to see if some preditors could be inferred as having better impact over mpg.
We'll then conclude with an `ANOVA` comparison of these two models with our initial model.1 that only looked at `mpg` vs. `am`.
```{r, results='hide'}
# All predictors model
model.all <- lm(mpg ~ ., data = mtcars) 
# Best model selection with stepwise
model.best <- step(model.all, direction = "both")
summary(model.best)
```

```{r}
printCoefmat(summary(model.best)$coefficients)
summary(model.best)$r.squared
```
From the stepwise model resulting `model.best`, we can see that a combination of 4 predictors: cylinder `cyl`, horse power `hp`, weight `wt` and transmission `am` explain about `r summary(model.best)$r.squared` (with an adjusted r squared of ~84%) of the mpg variance. We can see the p-value for `cyl`, horse power `hp`, weight `wt` are quite significant in their contribution to the accuracy of the model especially if compared to transmission `am`. 

Looking further at the *Residuals* (see point D in the appendix), we can see from plotting the model.best that:

- The residuals vs. fitted plot outlines data points that seem to be randomly scattered which indicates independence.
- The Normal Q-Q plot consists of data points that mostly fall on the regression line indicating a normal distribution.

```{r, results='hide'}
# Anova
anova(model.1, model.best, model.all)
```
In addition an ANOVA analysis of the variance (see point D in the appendix) with all models shows, that model.best has a significant p-value of `1.476e-05` which validates that additional predictors do have an impact on mpg and contribute to the accuracy of the overall model.

##### Conclusion
Manual transmission does have an increase of overall mpg performance of 1.80 mpg efficiency over automatic transmission, but only when including additional predictor variables to the model that are: cylinder number, horse power and car weight. The variable transmission alone is not sufficient to explain mpg efficiency.


\newpage

### Appendix
To see all R Code please find the [Rmd file](https://github.com/minitelle/RegMod/blob/master/motorstrend.Rmd).

#####A. List of variables used in analysis:

- am : transmission type: automatic/manual
- mpg: miles per gallon
- wt: car weight /1000lb
- cyl: engine cylinder
- hp: engine horse power

Source: https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html

###### Sample Data
```{r}
data("mtcars")
head(mtcars, n=5)
# Pais Graph to detect other predictors besides am
pairs(mtcars, panel=panel.smooth, main="Pair Graph of Motor Trend Car Road Tests")
```

####B. Figure 1 - MPG by Transmission type
```{r, echo = FALSE}
boxplot(mpg~am,data=mtcars, col=c("turquoise","tomato"),
        xlab="Fig 1. Transmission Type : 0-Automatic, 1-Manual",
        ylab="Mileage mpg",
        main="MPG by Trasmission type")
```

#####C. Residual plot for model.best (mpg ~ cyl + hp + wt + am)
```{r}
par(mfrow = c(2, 2)); plot(model.best)
```

#####D. ANOVA comparing (mpg ~am) vs. (mpg ~ cyl + hp + wt + am) vs. (mpg ~ .)
```{r}
anova(model.1, model.best, model.all)
```
